Particle Smoothing Algorithms for Variable Rate Models

### 2012-04-25 full_finaldraft_VRsmoothing_withSJGcomments.tex
Numerous small changes.
More edits. Added some bits about space of theta. About to make \theta_n into \theta_n^-

In a variable rate model, the set of changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are set out thoroughly in \cite{Jacobsen2006}. Conditional upon this MPP the state evolves according to some benign dynamics. In \cite{Godsill2007,Whiteley2011}, the conditional state evolution is treated as deterministic, while in \cite{Godsill2007a,Christensen2012} a conditionally linear-Gaussian state model is considered.

@@

The VRPF allows the changepoint sequence -- and hence the current state -- to be estimated sequentially as observations are received. However, estimates can often be improved later once further observations have been made. In this paper, we address the problem of smoothing in variable rate models, i.e. the estimation of the changepoint sequence and latent state given all the observations. This is achieved in this paper with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}. Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.

@@

This density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$.

In the manner of \cite{Whiteley2011}, a survivor function is defined as the probability that no new changepoint occurs before a given time,
%
\begin{IEEEeqnarray}{rCl}
 S(\tau_k, u_k, t) &=& P(\tau_{k+1}>t|\tau_k, u_k) \nonumber \\
              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}, u_k) d\xi     .
\end{IEEEeqnarray}

It is now possible to write down a prior for the changepoint sequence, where we use the convention that $\tau_0 = 0$. The existence of such a density for a MPP is addressed in \cite{Jacobsen2006}.

@@

This comprises a density term for each changepoint in the sequence and a survivor function term accounting for the probability that no other changepoints occur within the time interval.

\subsection{Conditionally Linear-Gaussian Models}

@@

\subsection{Conditionally Deterministic Models}

The second class of variable rate models for consideration is those in which the state is completely specified by the changepoint sequence, with no additional random components. Such a process is commonly referred to as ``piecewise-deterministic'', as the latent state follows a deterministic path between changepoints. In this case, it is not necessary to discretise the state -- it may be kept as a continuous variable.

In general, the state dynamics will be governed by a differential equation which may depend on the entire changepoint sequence. Here we assume that only the most recent changepoint determines the future state trajectory,
%
\begin{IEEEeqnarray}{rCl}
 dx(t) & = & \mathrm{f}(x(t), \tau_{K(t)}, u_{K(t)})     .
\end{IEEEeqnarray}

@@

By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time. Similarly, by choosing $t=t_n$, the state at the observation times may be calculated --- these latter points will be denoted $\hat{x}_n$. To complete the model, a probabilistic measurement model must be devised for the observation process, $p(y_n|\hat{x}_n)$.

@@

The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a,Whiteley2011}. The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$. This distribution may be expanded using Bayes' rule,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ p(\theta_{n}|y_{1:n}) \propto p(y_n|\theta_{n}, y_{1:n-1}) } \nonumber \\
 \qquad & & \times p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}     .
\end{IEEEeqnarray}

@@

where $\delta_x(X)$ is a dirac probability mass at $X=x$.

@@

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}$ to create a proposal for $\theta_n$. Finally, the particle is weighted according to the ratio of the target and proposal densities,
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) }{ q(\theta_{n-1}) q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}     .
\end{IEEEeqnarray}

@@

For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}). This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) until one falls after the current time, $t_n$. This final future changepoint is discarded. (This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula, % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}     .
\end{IEEEeqnarray}

@@

For a conditionally linear-Gaussian model, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter. Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations. It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
%
\begin{IEEEeqnarray}{rCl}
 p(x_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
 p(x_n|\theta_{n}, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
 p(y_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
\end{IEEEeqnarray}

@@

When a conditionally deterministic model is used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}). Thus, the required likelihood term is simply given by,
%
\begin{IEEEeqnarray}{rCl}
 p(y_n|\theta_{n}, y_{1:n-1}) & = & p(y_n|\hat{x}_n)     .
\end{IEEEeqnarray}

This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.

@@

At time $t_n$, the target distribution may be factorised as follows,
%
\begin{IEEEeqnarray}{rCl}
 p(\theta|y_{1:N}) = p(\theta_{n}^{+}|y_{1:N}) p(\theta_{n}|\theta_{n}^{+}, y_{1:N})     .
\end{IEEEeqnarray}

This suggests a sequential strategy for particle smoothing. Starting with a low diversity set of particles from $p(\theta|y_{1:N})$, we first marginalise $\theta_{n}$ by simply discarding the changepoints before $t_n$. New values of $\theta_{n}$ are then sampled from the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$, which is approximated using the filter particles, and the past and future sequences are concatenated. By iterating backwards from $n = N \dots 1$, a diverse set of particles approximating $p(\theta|y_{1:N})$ will be generated.

@@

Finally, the particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}|y_{1:n})$,
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(\theta_{n}|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n}) \label{eq:rb-vrps_back_cond}     ,
\end{IEEEeqnarray}

@@

The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$. This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}. Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF},
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \label{eq:backward_kf_predict_start} \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict_stop} \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \label{eq:backward_kf_update_start} \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
 \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}     .
\end{IEEEeqnarray}

Exact methods for initialising this recursion are discussed in appendix~\ref{app:init_backwards_KF}, and also in \cite{Kitagawa1994}. However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} = m_N^{(i)}$ and $\tilde{P}_N^{(i)} = P_N^{(i)}$.

Substituting the Gaussian densities into (\ref{eq:rb-vrps_back_cond_weight}) and performing the integral yields the following backwards conditional weights,
%
\begin{equation}
 \tilde{w}_n \propto w_n p(\theta_{n}^+|\theta_{n}^{(i)}) \mathcal{N}(\tilde{m}_n^-|m_n, \tilde{P}_n^- + P_n)     .
\label{eq:rb-vrps_back_cond_weight2}
\end{equation}

Using these weights, a sample of $\theta_{n}$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing one step of the smoothing algorithm. Once sampling has progressed backwards from $n=N \dots 1$, a particle from the smoothing distribution will have been generated. This procedure may then be repeated until sufficient particles have been obtained. The algorithm is summarised in Fig.~\ref{alg:RBVRPS}.

@@

In a conditionally deterministic system, the entire continuous-time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$. When the changepoint sequence is split into past, $\theta_n$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future. This is undesirable in practice, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.

@@

Again, this suggests a backwards sequential sampling algorithm for smoothing. The second term may be expanded as,
%
\begin{IEEEeqnarray}{rCl}
p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N}) & \propto & p(y_{n+1:N}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}) \nonumber \\
& & \times p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}|y_{1:n})    .
\end{IEEEeqnarray}

This presents a problem. We cannot directly substitute the particle approximation for the filtering distribution because the transition density term for the augmented sequences is given by,
%
\begin{IEEEeqnarray}{rCl}
p(\tilde{\theta}_{n}^+ | \tilde{\theta}_{n}) & \propto & p(\tau_{K_n+1}, u_{K_n+1}| \tau_{K_n}, u_{K_n}) \nonumber \\
 & & \times \delta_{f(x_{K_n}, u_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1}),
\end{IEEEeqnarray}

which will be $0$ for all but one unique filtering particle. Intuitively, we are trying to join two deterministic state trajectories together, but they do not meet up in the middle.

The solution to this problem is provided by an idea from the Sequential Monte Carlo (SMC) samplers of \cite{DelMoral2006}. Rather than sampling $p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})$ directly, a value of $\tilde{\theta}_{n}$ is proposed from the filtering particles, after which a replacement, $u_{K_n}'$, is proposed for the existing final changepoint parameter, with a value chosen such that the past and future paths meet up. The modified sequence is denoted $\tilde{\theta}_{n}'$. Such a proposal density may be written as,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}|y_{1:n}) q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})} \nonumber \\
 \qquad & \propto & p(y_{1:n}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}) q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_proposal}     .
\end{IEEEeqnarray}

The target distribution is augmented by introduction of an artificial density to cover the new changepoint sequence and the discarded parameter.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N}) L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) } \nonumber \\
 & \propto & p(y_{1:N}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^+) p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') p(\tilde{\theta}_{n}') L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N})  \label{eq:pd-vrps_target}     .
\end{IEEEeqnarray}

@@

The proposal density $q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$ is constructed so as to be non-zero only where $p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}')$ is non-zero. This could require it to be a delta function if there is only one feasible value. The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is simply as a uniform distribution so that it cancels out in the MH acceptance probability. The remaining terms are likelihoods and parameter transition density terms, defined by the system equations.

@@

The smoothers and the final step of the filters each produce a particle approximation to the changepoint sequence posterior, $p(\theta|y_{1:N})$. Calculating a single estimator from these is no simple task. Each particle consists of a list of changepoint times and parameters, and there is no obvious way to take an average over the set to produce a single estimate. Furthermore, if a best estimate could be calculated (e.g. by finding the maximum a posteriori sequence), it is not trivial to define a distance metric between the estimated and true sequences to assess relative accuracy.

@@

The Rao-Blackwellised VRPS algorithm was tested on the financial time series model of \cite{Godsill2007a,Christensen2012}, in which prices of an asset are treated as noisy observations of a latent state, which evolves according to a drift-diffusion with occasional jumps.

@@

The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:finance_unique_particles}. This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution.

>>>>

In a variable rate setting, changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are explored comprehensively in \cite{Jacobsen2006}. Conditional upon this MPP the state evolves according to some benign dynamics. In \cite{Godsill2007,Whiteley2011}, the conditional state evolution is treated as deterministic, while in \cite{Godsill2007a,Christensen2012} a conditionally linear-Gaussian state model is considered.

@@

The VRPF allows the changepoint sequence -- and hence the current state -- to be estimated sequentially as observations are received. However, estimates can often be improved later once further observations have been made. In this paper, we address the problem of smoothing in variable rate models, i.e. the estimation of the changepoint sequence and latent state given all the observations. We achieve this with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}. Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.

@@

This density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$. As in \cite{Whiteley2011}, a survivor function is defined as the probability that no new changepoint occurs before a given time,
%
\begin{IEEEeqnarray}{rCl}
 S(\tau_k, u_k, t) &=& P(\tau_{k+1}>t|\tau_k, u_k) \nonumber \\
              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}, u_k) d\xi     .
\end{IEEEeqnarray}

It is now possible to write down a prior for the changepoint sequence. This comprises a density term for each changepoint in the sequence and a survivor function term accounting for the probability that no additional changepoints occur within the time interval. We use the convention that $\tau_0 = 0$.

@@

The existence of such a density for a MPP is addressed in \cite{Jacobsen2006}. The number of changepoints, $K_n$, is not fixed. If $u_k$ varies over $\mathcal{U}$, and $\tau_k$ varies over the real line $\mathbb{R}$, then the space spanned by $\theta_n$ is the union,
%
\begin{IEEEeqnarray}{rCl}
 \Theta & = & \bigcup_k \Theta_k \label{eq:theta_space} \\
 \Theta_k & = & \{k\} \times \mathbb{R}^k \times \mathcal{U}^k     .
\end{IEEEeqnarray}

This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo algorithms \cite{Green1995}.% or in multiple target tracking scenarios \cite{Vo2005}.

\subsection{Conditionally Linear-Gaussian Models} \label{sec:clg_models}

@@

\subsection{Conditionally Deterministic Models} \label{sec:cd_models}

Next we examine the class of variable rate models in which the state is completely specified by the changepoint sequence, with no additional random components. Such a process is commonly referred to as ``piecewise-deterministic'', as the latent state follows a deterministic path between changepoints. In this case, it is not necessary to discretise the state --- it may be kept as a continuous variable.

In general, the state dynamics will be governed by a differential equation which may depend on the entire changepoint sequence. Here we assume that only the most recent changepoint determines the future state trajectory,
%
\begin{IEEEeqnarray}{rCl}
 dx(t) & = & \varphi(x(t), \tau_{K(t)}, u_{K(t)}) dt     .
\end{IEEEeqnarray}

@@

By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time. Similarly, by choosing $t=t_n$, the state at the observation times may be evaluated --- these latter points will be denoted $\hat{x}_n$. To complete the framework, a probabilistic measurement model must be devised for the observation process, $p(y_n|\hat{x}_n)$.

@@

The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a,Whiteley2011}. The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$, which is defined on the same space as the prior density (\ref{eq:theta_space}). This distribution may be expanded using Bayes' rule,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ p(\theta_{n}|y_{1:n}) \propto p(y_n|\theta_{n}, y_{1:n-1}) } \nonumber \\
 \qquad & & \times p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}     .
\end{IEEEeqnarray}

@@

where $\delta_x(X)$ is a unit probability mass at $X=x$.

@@

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}^{(i)}$ to create a proposal for $\theta_n^{(i)}$. Finally, the particle is weighted according to the ratio of the target and proposal densities,
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}^{(i)}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n^{(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}) p(\theta_{n-1}^{(i)}|y_{1:n-1}) }{ q(\theta_{n-1}^{(i)}) q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n^{(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}) }{ q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}     .
\end{IEEEeqnarray}

@@

For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}). This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) until one falls after the current time, $t_n$. This final future changepoint is discarded. (This process can be thought of as sampling an entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula, % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}     .
\end{IEEEeqnarray}

@@

For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter. Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations. It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
%
\begin{IEEEeqnarray}{rCl}
 p(x_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
 p(x_n|\theta_{n}, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
 p(y_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
\end{IEEEeqnarray}

@@

When the conditionally deterministic models of section~\ref{sec:cd_models} are used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}). Thus, the required likelihood term is simply given by,
%
\begin{IEEEeqnarray}{rCl}
 p(y_n|\theta_{n}, y_{1:n-1}) & = & p(y_n|\hat{x}_n)     .
\end{IEEEeqnarray}

This leads to a piecewise-deterministic variable rate particle filter, as summarised in Fig.~\ref{alg:VRPF}.

@@

For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the target distribution may be factorised as follows at time $t_n$,
%
\begin{IEEEeqnarray}{rCl}
 p(\theta|y_{1:N}) = p(\theta_{n}^{+}|y_{1:N}) p(\theta_{n}|\theta_{n}^{+}, y_{1:N})     .
\end{IEEEeqnarray}

This suggests a sequential strategy for particle smoothing. Starting with a low diversity set of particles from $p(\theta|y_{1:N})$, we first marginalise $\theta_{n}$ by simply discarding the changepoints before $t_n$. New values of $\theta_{n}$ are then sampled from the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$, which is approximated using the filter particles, and the past and future sequences are concatenated. By iterating backwards over $n = N \dots 1$, a diverse set of particles approximating $p(\theta|y_{1:N})$ will be generated.

@@

The particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}|y_{1:n})$, yielding,
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(\theta_{n}|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n}) \label{eq:rb-vrps_back_cond}     ,
\end{IEEEeqnarray}

@@

The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$. This may be regarded as an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}. Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF},
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \label{eq:backward_kf_predict_start} \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict_stop} \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \label{eq:backward_kf_update_start} \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
 \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}     .
\end{IEEEeqnarray}

Exact methods for initialising this recursion are discussed in appendix~\ref{app:init_backwards_KF}, and also in \cite{Kitagawa1994}. However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} \gets m_N^{(i)}$ and $\tilde{P}_N^{(i)} \gets P_N^{(i)}$.

Substituting the Gaussian densities into (\ref{eq:rb-vrps_back_cond_weight}) and integrating yields the following backwards conditional weights,
%
\begin{equation}
 \tilde{w}_n \propto w_n p(\theta_{n}^+|\theta_{n}^{(i)}) \mathcal{N}(\tilde{m}_n^-|m_n, \tilde{P}_n^- + P_n)     .
\label{eq:rb-vrps_back_cond_weight2}
\end{equation}

Using these weights, a sample of $\theta_{n}$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing one step of the smoothing algorithm. Once sampling has progressed backwards over $n=N \dots 1$, a particle from the smoothing distribution will have been generated. This procedure may then be repeated until sufficient particles have been obtained. The algorithm is summarised in Fig.~\ref{alg:RBVRPS}.

@@

In the conditionally deterministic models of section~\ref{sec:cd_models}, the entire continuous-time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$. When the changepoint sequence is split into past, $\theta_n$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future. This is undesirable in practice, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.

@@

As before, this suggests a backwards sequential sampling algorithm for smoothing. The second term may be expanded as,
%
\begin{IEEEeqnarray}{rCl}
p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N}) & \propto & p(y_{n+1:N}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}) \nonumber \\
& & \times p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}|y_{1:n})    .
\end{IEEEeqnarray}

A difficulty is now encountered. We cannot directly substitute the particle approximation for the filtering distribution because the transition density term for the augmented sequences is given by,
%
\begin{IEEEeqnarray}{rCl}
p(\tilde{\theta}_{n}^+ | \tilde{\theta}_{n}) & \propto & p(\tau_{K_n+1}, u_{K_n+1}| \tau_{K_n}, u_{K_n}) \nonumber \\
 & & \times \delta_{f(x_{K_n}, u_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1}),
\end{IEEEeqnarray}

which will be $0$ for all but one unique filtering particle, the one from which the next selected future changepoint originated. Intuitively, we are trying to join two deterministic state trajectories together, but they do not meet up in the middle.

The solution to this problem is provided by an idea from the Sequential Monte Carlo (SMC) samplers of \cite{DelMoral2006}. Rather than sampling $p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})$ directly, an augmented distribution is targeted. A value of $\tilde{\theta}_{n}$ can be proposed from the filtering particles, after which a replacement, $u_{K_n}'$, is proposed for the existing final changepoint parameter, with a value chosen such that the past and future paths meet up. The modified sequence is denoted $\tilde{\theta}_{n}'$. Such a proposal density may be written as,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}|y_{1:n}) q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})} \nonumber \\
 \qquad & \propto & p(y_{1:n}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}) q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_proposal}     .
\end{IEEEeqnarray}

The augmented target distribution over the the new sequence, $\tilde{\theta}_{n}'$, and the discarded parameter, $u_{K_n}$, is constructed by introducing an artificial density,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N}) L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) } \nonumber \\
 & \propto & p(y_{1:N}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^+) p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') p(\tilde{\theta}_{n}') L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N})  \label{eq:pd-vrps_target}     .
\end{IEEEeqnarray}

@@

The proposal density $q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$ is constructed so as to be non-zero only where $p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}')$ is non-zero. This could require it to be a delta function if there is only one feasible value. The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is a uniform distribution over the set of parameter values, $\{u_{K_n}^{(i)}\}$, which appear in the filtering approximation. With this choice, the artificial density terms will always cancel out in the MH acceptance probability. The remaining terms are likelihoods and parameter transition density terms, defined by the system equations.

@@

The smoothers and the final step of the filters each produce a particle approximation to the changepoint sequence posterior, $p(\theta|y_{1:N})$. Calculating a single estimator from these is no simple task. Each particle consists of a list of changepoint times and parameters, and there is no obvious way to take an average over the set to produce a single estimate. Furthermore, if a best estimate could be calculated (e.g. by finding the maximum a posteriori sequence), it is not trivial to define a distance metric between the estimated and true sequences to assess relative accuracy. (One possible choice of metric has been considered for this comparison, the optimal sub-pattern assignment (OSPA) distance of \cite{Schuhmacher2008}. However, the relative performance of the algorithms is liable to vary significantly with the parameters of the metric.)

@@

The Rao-Blackwellised VRPS algorithm was tested on the financial time series model of \cite{Christensen2012}, in which prices of an asset are treated as noisy observations of a latent state, which evolves according to a drift-diffusion with occasional jumps.

@@

The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:finance_unique_particles}. This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint and state distributions.
